On the surprising effectiveness of a simple matrix exponential derivative approximation, with application to global SARS-CoV-2

### 2023-10-25 manuscript/pnas_revision.tex
finished captions

\begin{figure}[!ht]
	\centering
\includegraphics[width=\linewidth]{lowDimVis.pdf}
\vspace{-2em}
\caption{All methods agree.}\label{fig:lowDim}
\end{figure}

\begin{figure*}[!t]
	\centering
	\includegraphics[width=\linewidth]{truthMeanFig.pdf}
	\caption{Frobenius norm errors obtained by first-order approximation $te^{t\QQ}\JJ$ and by affine-corrected first-order approximation
		$te^{t\QQ}\JJ - t(\II -\QQ^+ \QQ) \JJ \QQ \QQ^{+}$
		(\cref{thm:bride:of:frankinstein}) under increasingly relaxed
		assumptions.  Within each assumption set, we average over 20
		independent Monte Carlo simulations of random generator matrices
		for each dimension. Plot \textbf{a} corresponds to asymmetric
		generators with off-diagonal elements having independent and
		identically distributed (iid) standard exponential random
		variables that correspond to the sub-exponential distribution
		hypothesis. Plot \textbf{b} drops the identical distribution
		assumption by allowing each row and column of the generator
		matrix to additively contribute its own mean---itself given by a
		standard exponential---to its corresponding exponentially
		distributed entries. Plot \textbf{c} features rate
		matrices with iid Cauchy entries truncated to be
		positive. Empirically, the results of
		\cref{cor:rig:det:rand:wrapup} extend beyond the symmetric,
		iid and sub-exponential hypotheses, suggesting scope of future work.}
	\label{fig:truthMean}
\end{figure*}

>>>>

\begin{figure}[!ht]
	\centering
\includegraphics[width=\linewidth]{lowDimVis.pdf}
\vspace{-2em}
\caption{Posterior density plots for surrogate-trajectory Hamiltonian Monte Carlo (HMC) (\cref{sec:Sur:trad:HMC}) using the naive approximate derivative, the corrected approximation \eqref{eq:better:help:on:sale} and the exact matrix exponential derivative for elements of a $5\times 5$ generator matrix. The near perfect overlap reflects the fact that each algorithm's transition kernel leaves the posterior distribution invariant \cite{glatt2020accept}.  To generate data, we randomly draw standard normal generator entries once and simulate 20 independent initial/final position pairs from a CTMC with time interval $t=1$.  We show the true value in red and negate diagonal elements to simplify presentation.  }\label{fig:lowDim}
\end{figure}

\begin{figure*}[!t]
	\centering
	\includegraphics[width=\linewidth]{truthMeanFig.pdf}
	\caption{Posterior means versus truth for CTMC generator matrix elements under differing sparsity regimes and different dimensionalities $d$ holding the number of observations fixed at 300.  To affect sparsity levels, we generate generator matrix entries according to the Bayesian bridge distribution \cite{polson2014bayesian} with different exponents ($\alpha \in \{1,1/2,1/4\}$), normalizing by the largest absolute values to ease comparison. Smaller $\alpha$ values encode more peaked distributions with heavier tails and thus enforce greater sparsity.   Plot \textbf{c} reflects the fact that the Bayesian bridge prior with exponent $\alpha=1/4$ helps identify non-null parameters in small sample contexts \cite{magee2023random}.  With this intuition in mind, we specify such a prior on generator random effects in \cref{sec:sars:cov:app}.}
	\label{fig:truthMean}
\end{figure*}
